---
title: "Data Import"
author: "Jarred Robidoux"
date: "2023-02-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# __Data Import__
Working with data provided by R packages is a great way to learn the tools of data 
science, but at some point you want to stop learning and start working with your 
own data. In this walkthrough, you'll learn how to read plain-text rectangular files
into R. 

# __Prerequisites__
In this chapter, you'll learn how to load flat files in R with the readr package
which is part of the core tidyverse
```{r}
library(tidyverse)
```

# __Getting Started__
Most of readr's functions are concerned with turning flat files into data frames

**read_csv()** reads comman delimted files, **read_csv2()** reads semicolon separated
files (common in countries where , is used as the decimal place), **read_tsv()** reads
tab delimited files, and **read_delim()** reads in files with any delimiter.


**read_fwf()** reads fixed width files. You can specify fields either by their widths
with **fwf_widths()** or their position with **fwf_positions()**. **read_table()** reads
a common variation of fixed width files where columns are separated by white space

**read_log()** reads Apache style log files. (But also check out webreadr which is 
built on top of **read_log()** and provides many more helpful tools)

The first argument to **read_csv()** is the most important: it's the path to the file
to read
```{r}
airtravel <- read_csv("airtravel.csv", col_names = T)
```

You can also supply an inline csv file. This is useful for experimenting with readr
and for creating reproducible examples to share with others. 
```{r}
read_csv("a,b,c
         1,2,3
         4,5,6")
```

Sometimes there are a few lines of metadata at the top of the file. You can use 
**skip** = n to skip the first n lines; or use **comment** = "#" to drop all lines that start 
with (e.g.) #.
```{r}
read_csv("The first line of metadata
         The second line of metadata
         x,y,z
         1,2,3", skip = 2)
```

The data also might not have column names. You can use **col_names** = FALSE to tell
**read_csv()** not to treat the first row as headings, and instead label them sequentially from X1 to Xn.
```{r}
read_csv("1,2,3\n4,5,6", col_names = FALSE)
```

Alternatively you can pass **col_names()** a character vector which will be used as the column names. 
```{r}
read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z"))
```

Another option that commonly needs tweaking is **na**: this specifies the value (or values) that are used to represent missing values in your file. 
```{r}
read_csv("a,b,c\n1,2,.", na = ".")
```

# __Exercises__ 
1. What function would you use to read a file where fields were separated with "|"
```{r}
read_delim(file, delim = "|")
```

2. Apart from **file**, **skip**, and **comment**, what other arguments do read_csv() and read_tsv() have in common?
```{r}
?read_csv
```

Some common arguments include; na =, delim =, col_names = 

3. What are the most important arguments to read_fwf()
```{r}
?read_fwf
```

The most important arguments to read_fwf() are file, col_positions, col_types, and col_select

4. Sometimes strings in a CSV file contain commas. To prevent them from causing problems they need to be surrounded by a quoting character, like **"** or **'**. By default, read_csv() assumes that the quoting character will be ". What argument to read_csv() do you need to specify to read the following text into a data frame? 
```{r}
read_csv("x,y\n'a,b'", quote = "'")
```

5. Identify what is wrong with each of the following inline CSV files. What happens when you run the code? 
```{r}
read_csv("a,b\n1,2,3\n4,5,6")
```

