---
title: "Parsing Vectors and Files"
author: "Jarred Robidoux"
date: "2023-02-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
tinytex::install_tinytex(force = TRUE)
```

# __Parsing a Vector__
Let's talk about the **parse**() functions. These functions take a character vector 
and return a more specialized vector like a logical, integer, or date. 

```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
```
```{r}
str(parse_integer(c("1", "2", "3")))
```
```{r}
str(parse_date(c("2010-10-01", "1979-10-14")))
```
Like all function in the tidyverse, the **parse**() functions are uniform: the first argument is a character vector to parse, and the na argument specifies which strings should be treated as missing:
```{r}
parse_integer(c("1", "231", ".", "456"), na = ".")
```
If parsing fails, you'll get a warning: 
```{r}
x <- parse_integer(c("123", "345", "abc", "123.45"))
```

If there are many parsing failures, you can use **problems()** to get the complete 
set. This returns a tibble, which you can then manipulate with dplyr

```{r}
problems(x)
```
# __Types of Parse()__
1. **parse_logical()** and **parse_integer()** parse logicals and integers respectively. There's basically nothing that can go wrong with these parsers.

2. **parse_double()** is a strict numeric parser, and **parse_number()** is a flexible
numeric parser. These are more complicated than you might expect because different parts of the world write numbers in different ways. 

3. **parse_character()** seems so simple that it shouldn't be necessary. But one complication makes it quite important: character encoding.

4. **parse_factor()** create factors, the data structure that R uses to represent categorical variables with fixed and known values. 

5. **parse_datetime()**, **parse_date()**, and **parse_time()** allow you to parse various date & time specifications. These are the most complicated because there are so many different ways of writing dates. 

# __Numbers__
It seems like it should be straightforward to parse a number, but three problems make it tricky:

1. People write numbers differently in different parts of the world

2. Numbers are often surrounded by other characters that provide some context, like $1000 or 10%

3. Numbers often contains "grouping" characters to make them easier to read, like "1,000,000", and these grouping character vary around the world. 

To address these problems we can use the notion of a "locale", an object that specifies parsing options that differ from place to place
```{r}
parse_double("1.23")
```
```{r}
parse_double("1,23", locale = locale(decimal_mark = ","))
```

```{r}
parse_number("$100")
```

```{r}
parse_number("20%")
```

```{r}
parse_number("It cose $123.45")
```
The final problem is addressed by the combination of **parse_number()** and the locale as **parse_number()** will ignore the "grouping mark"
```{r}
parse_number("$123,456,789", locale = locale(grouping_mark = ","))
```

# __Strings__
Similar to numbers, you can use **locale** to turn character strings into other languages
```{r}
x2 <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"
parse_character(x2, locale = locale(encoding = "Shift-JIS"))
```

# __Factors__
Within the **parse_factor()** function, we can use **levels=** to signify how exactly to sort the strings as factors.
```{r}
fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "banana"), levels = fruit)
```
# __Dates, date-times, and times__
You pick between three parsers depending on whether you want a date (the number of days since 1970-01-01), a date-time (the number of seconds since mignight 1970-01-01), or a time (the number of seconds since midnight). When called without any additional arguments.
```{r}
parse_datetime("2010-10-01T2010")
```

**parse_date()** expects a four digit year, a **-** or **/**, the month, a **-** or **/**, then the day:
```{r}
parse_date("2010-10-01")
```
**parse_time()** expects the hour, **:**, minutes, optionally **:** and seconds, and an optional am/pm specifer.
```{r}
parse_time("01:10am")
```
```{r}
parse_time("20:10:01")
```

If these defaults don't work for your data you can supply your own date-time format, built up of the following pieces:
*Year*

**%Y** (4digits)
**%y** (2 digits)

*Month*
**%m** (2 digits)
**%b** (abbreviated name)
**B%** (full name, "January")

*Day*
**%d** (2 digits)
**%e** (optional leading space)

*Time*
**%H** (0-23 hour)
**%I** (0-12, must be used with %p)
**%p** (AM/PM indicator)
**%M** (minutes)
**%S** (integer seconds)
**%OS** (Real seconds)
**%Z** (Time Zone)
**%z** (as offset from UTC, eg +0800)

Examples
```{r}
parse_date("01/02/15", "%m/%d/%y")
```
```{r}
parse_date("01/02/15", "%d/%m/%y")
```

# __Parsing a File__
readr uses a heuristic to figure out the type of each column: it reads the first 1000 rows and uses some (moderately conservative) heuristics to figure out the type of each column. 
You can emulate this process with a character vector using **guess_parser()**, which returns readr's best guess, and **parse_guess()** which uses that guess to parse a column. 
```{r}
guess_parser("2010-10-01")
guess_parser(c("TRUE", "FALSE", "T"))

str(parse_guess("2010-10-01"))
```
# __Writing to a File__
readr also comes with two useful functions for writing data back to disk: **write_csv()** and **write_tsv()**. 
```{r}
challenge <- read_csv(readr_example("challenge.csv"))
write_csv(challenge, "challenge.csv")
```


